Compositional Quantitative Resource Analysis of C Programs

### 2014-06-18 paper/main.tex
more work on section 2

XXX add up and down XXX

@@

@@

For the sake of simplicity, our quantitative program logic
applies to the Clight language.  This language is the most
abstract intermediate language used by CompCert.  It is very
similar to vanilla C but the expressions are free of
side-effects and there is only one looping construct.
%
\begin{align*}
S &:= \code{assert}~E
\mid \code{skip}
\mid \code{break}
\mid \code{return}~E
\mid x \gets E
\mid x \gets f(x^*)
\\
& \mid \code{loop}~S
\mid \code{if}(E)~S~\code{else}~S
\mid S;S
\end{align*}
%
We added the builtin primitive $\code{assert}$ that checks
if a condition is true before continuing the program
execution.  This will be used to explicit invariants to the
automatic analysis later.

@@

The operational semantics uses a finite map $\state$ from
variable names to integers.  This map is represented as
an associative list to model the semantics of the stack of
our language. It can be extended with new bindings
(hiding older bindings with the same name), shrinked by
forgetting a list of bindings, and updated by changing one
binding.  We note $(\state, x_1 \mapsto v_1, \dots, x_n \mapsto v_n)$
an extended heap, $\state~\backslash~(x_1, \dots, x_n)$ a
state where bindings of $x_1, \dots, x_n$ were removed and
$\state[x \mapsto v]$ an updated state.

To formally describe the cost of operation of a given
program, our semantics tracks resource consumption while
executing.  The semantics is oblivious to the resource of
interest for the user of our system.  We acheive this
independence by parameterizing the semantics by a resource
\emph{metric}, it is a tuple of rational numbers with the
following form
$$
(
M_a, M_s, M_b, M_r, M_u, M_f,
M_l,
M_c^1, M_c^2,
M_q
).
$$
Each of these rational numbers indicates the amount of resource
consumed by an operation.  If one of these numbers is negative,
it means that resources are released by the program.

The general form of an evaluation judgement is
$(\state, S) \Downarrow_{(q,q')} O$.  It is similar to the judgement
presented in the introducing section appart from the fact that
two resource numbers are tracked instead of one.  Intuitively
$(q, q')$ means that $q$ resources units are necessary to execute
the program and that $q'$ units are available after its execution.
For example, $\code{alloc}(3)$ consumes 3 heap cells, so
the pair resulting from its evaluation would be $(3, 0)$; for
$\code{free}(3)$, the pair would be $(0, 3)$.  When these
statements are sequenced with \code{alloc} first, then \code{free},
the evaluation yields $(3, 3)$.  These resource pairs can
be composed with the following operation
$$
(q_1, q'_1) \cdot (q_2, q'_2) =
\left\{
\begin{array}{ll}
(q_1, q'_2 + q'_1 - q_2) & \mbox{if } q'_1 \ge q_2 \\
(q_1 + q_2 - q'_1, q'_2) & \mbox{if } q'_1 < q_2
\end{array}
\right.
.
$$
This operation can be seen as the sequential composition of
two resource consumptions, note that it is not a commutative
operation, the same way resource freeing and allocation do not
commute.
With $(0, 0)$ as a neutral element, the above composition defines
a monoid structure on $\mathbb Q_0^+ \times \mathbb Q_0^+$.
It is possible to inject $q \in \mathbb Q$ in this monoid by
turning it into $(q, 0)$ if $q \ge 0$ and $(0, -q)$ otherwise.
This injection is used implicitely in the semantics.

\Rule{S:Abort}
{ }
{ (\state, S) \Downarrow_{(0,0)} \circ }

\and \Rule{S:Assert}
{ {\rm istrue}~\llbracket e \rrbracket_\state}
{ (\state, \code{assert}~e) \Downarrow_{M_a} \code{seq}~\state }

\and \Rule{S:Skip}
{ }
{ (\state, \code{skip}) \Downarrow_{M_s} \code{seq}~\state }

\and \Rule{S:Break}
{ }
{ (\state, \code{break}) \Downarrow_{M_b} \code{brk}~\state }

\and \Rule{S:Return}
{ }
{ (\state, \code{return}~e) \Downarrow_{M_r}
  \code{ret}~\state~\llbracket e \rrbracket_\state
}

\and \Rule{S:Update}
{ }
{ (\state, x \gets e) \Downarrow_{M_u}
  \code{seq}~\state[x \mapsto \llbracket e \rrbracket_\state]
}

\and \Rule{S:Call}
{ \delta~f = (\vec y, S) \\
  ((\state, y_i \mapsto \state(x_i)), S)
    \Downarrow_{(q,q')} \code{ret}~\state'~v
}
{ (\state, r \gets f(\vec x)) \Downarrow_{M_f\cdot(q,q')}
  \code{seq}~(\state'~\backslash~(x_i))[r \mapsto v]
}

\and \Rule{S:LoopSeq}
{ (\state, S) \Downarrow_{(q_1, q_1')} \code{seq}~\state'
\\ (\state', \code{loop}~S) \Downarrow_{(q_2, q_2')} O }
{ (\state, \code{loop}~S) \Downarrow_{(q_1, q_1') \cdot M_l \cdot (q_2, q_2')} O }

\and \Rule{S:LoopBrk}
{ (\state, S) \Downarrow_{(q, q')} \code{brk}~\state' }
{ (\state, \code{loop}~S) \Downarrow_{(q, q')} \code{seq}~\state' }

\and \Rule{S:LoopAbort}
{ (\state, S) \Downarrow_{(q, q')} O
\\ \forall \state.~O \neq \code{seq}~\state \land O \neq \code{brk}~\state }
{ (\state, \code{loop}~S) \Downarrow_{(q, q')} O }

\and \Rule{S:IfThen}
{  (\state, S_1) \Downarrow_{(q, q')} O
\\ {\sf istrue}~\llbracket e \rrbracket_\state
}
{ (\state, \code{if}(e)~S_1~\code{else}~S_2)
  \Downarrow_{M_c^1 \cdot (q, q')} O
}

\and \Rule{S:IfElse}
{  (\state, S_2) \Downarrow_{(q, q')} O
\\ {\sf isfalse}~\llbracket e \rrbracket_\state
}
{ (\state, \code{if}(e)~S_1~\code{else}~S_2)
  \Downarrow_{M_c^2 \cdot (q, q')} O
}

\and \Rule{S:Seq}
{  (\state, S_1) \Downarrow_{(q_1, q'_1)} \code{seq}~\state'
\\ (\state', S_2) \Downarrow_{(q_2, q'_2)} O
}
{ (\state, S_1; S_2) \Downarrow_{(q_1, q'_1) \cdot M_q \cdot (q_2, q'_2)} O }

\and \Rule{S:SeqAbort}
{ (\state, S_1) \Downarrow_{(q,q')} O
\\ \forall \state.~O \neq \code{seq}~\state}
{ (\state, S_1; S_2) \Downarrow_{(q,q')} O }

@@

In the evaluation judgement $(\state, S) \Downarrow_{(q, q')} O$,
the outcome $O$ specifies how the statement $S$ terminated.
$$
O := \circ
\mid \code{seq}~\sigma
\mid \code{brk}~\sigma
\mid \code{ret}~\sigma~v
$$
The \code{seq} outcome is the simple fallthrough outcome, for example
a \code{skip} statement will terminate with this outcome without changing
the heap as pictured in the rule {\sc S:Skip}.  The \code{brk} and
\code{ret} outcomes describe a statement that terminated with a break or
return statement.  In the latter case, a return value is also part of the
outcome.  The use of these different outcomes can be explained by considering
the loop statement rules:  If the inner statement breaks
then the loop exits using the rule {\sc S:LoopBrk}, if the inner statement
falls through another iteration is run using the rule {\sc S:LoopSeq},
and all other cases are handled in the rule {\sc S:LoopAbort}.

The special outcome $\circ$ indicates an early
termination, it can be triggered at any point using the rule {\sc S:Abort}
and bubbles up through the execution context.
Sometimes there is no other choice than to use this latter rule, for
example when an assertion fails.  In that case the side condition of
{\sc S:Assert} is not met and the only rule that can possibly be applied
is {\sc S:Abort}.

The busy result is also useful to handle diverging programs with big-step
judgements.  In case a program does not terminate, it is impossible to
derive $(\state, S)\Downarrow_{(q, q')} O$ with $O \neq \circ$.  The only
possible way to conclude a big-step judgement is to use the {\sc S:Abort}
rule at some point in the computation.  One alternative to this technique
is to use coinductive semantics~\cite{Leroy-coind}, however, using the busy
result prevents duplication and allows to have only one soundness proof for
both unsafe and non-terminating programs.  Indeed, stating a result for
$(\sigma, S) \Downarrow_{(q,q')} \circ$ covers unsafe programs, diverging
programs, and partial computations at the same time.

@@

\begin{figure*}[t]
\begin{mathpar}

\Rule{L:Assert}
{ }
{ \Delta; B; R \vdash
\htriple
  { Q \implies \code{istrue}~\sem{e}_\state }
  { \code{assert}~e }
  { Q }
}

\and
\Rule{L:Skip}
{ }
{ \Delta; B; R \vdash
\htriple
  { M_s + Q }{ \code{skip} }{ Q }
}

\and
\Rule{L:Break}
{ }
{ \Delta; B; R \vdash
\htriple
  { M_b + B }{ \code{break} }{ Q }
}

\and
\Rule{L:Return}
{ }
{ \Delta; B; R \vdash
\htriple
  { M_r + R\,\sem{e}_\state }{ \code{return}~e }{ Q }
}

\and
\Rule{L:Update}
{ }
{ \Delta; B; R \vdash
\htriple
  { \lambda \state.\,M_u + Q\,\state[x \mapsto \sem{e}_\state] }{ x \gets e }{ Q }
}

\and
\Rule{L:Call}
{
  \Delta f = \forall z\,\vec v.(P_f\,z\,\vec v, Q_f\,z) \\
  P \models P_f \, y \, (\sigma(\vec x)) \land A \\
  \forall v.\, (Q_f \, y \, v \land A \models \lambda \state.\, Q \, \sigma[r \mapsto v])
}
{ \Delta; B; R \vdash
\htriple
  { M_f + P }
  { r \gets f(\vec x) }
  { Q }
}

\and
\Rule{L:Loop}
{ \Delta; Q; R \vdash
\htriple
  { I }{ S }{ I + M_l }
}
{ \Delta; B; R \vdash
\htriple
  { I }{ \code{loop}~S }{ Q }
}

\and
\Rule{L:Seq}
{ \Delta; B; R \vdash
\htriple{ P }{ S_1 }{ Q' + M_s }
\\
  \Delta; B; R \vdash
\htriple{ Q' }{ S_2 }{ Q }
}
{ \Delta; B; R \vdash
\htriple{ P }{ S_1;S_2 }{ Q }
}

\and
\Rule{L:If}
{ \Delta; B; R \vdash
\htriple
  { P \land \code{istrue}~\sem{e}_\state - M_c^1}
  { S_1 }{ Q }
\\
  \Delta; B; R \vdash
\htriple
  { P \land \code{isfalse}~\sem{e}_\state - M_c^2 }
  { S_2 }{ Q }
}
{ \Delta; B; R \vdash
\htriple
  { P }{ \code{if}(e)~S_1~\code{else}~S_2 }{ Q }
}

\and
\Rule{L:Weaken}
{
  P \models P' \\
  \Delta; B'; R' \vdash
  \htriple{ P' }{ S }{ Q' } \\
  Q' \models Q \\
  B' \models B \\
  \forall v.\, (R'\,v \models R\,v)
}
{ \Delta; B; R \vdash
\htriple{ P }{ S }{ Q }
}

\and
\Rule{L:Frame}
{ \Delta; B; R \vdash
\htriple{ P }{ S }{ Q } \\
  x \in \mathbb Q_0^+
}
{ \Delta; B + x; R + x \vdash
\htriple{ P + x }{ S }{ Q + x }
}

\and
\Rule{L:Extend}
{ \forall f\,P_f\,Q_f.\,
  \Delta'f = \forall z\,\vec v.(P_f\,z\,\vec v, Q_f\,z)
  \rightarrow
  \forall y\,\vec v.\,
  (\Delta \cup \Delta'; \bot; Q_f\,y \vdash
    \htriple{ P_f\,y\,\vec v }{ \delta f }{ \bot })
\\
  \Delta \cup \Delta'; B; R \vdash
  \htriple{ P }{ S }{ Q }
}
{ \Delta; B; R \vdash
\htriple{ P }{ S }{ Q }
}

\end{mathpar}

>>>>

Example \emph{t08} shows the ability of the analysis to handle
negative and non-negative numbers.  Note that there are no
restrictions on the signs of $z$ and $y$.  We also see that we can
accurately track the size change of $y$ in the first loop.
Furthermore, \emph{t08} shows that we do not handle the constants $1$
or $0$ in any special way.  In all examples you could replace $0$ and
$1$ with other constants like we did in the second loop and still
derive a tight bound.  The only information, that the analyzer needs
is $y \geq c$ before assigning $y = y - c$.

Example \emph{t27} shows how amortization can be used to handle tricky
nested loops.  The outer loop is iterated $|[n,0]|$ times.  In the
outer loop, we increment the variable $n$ till $n = 0$.  In each of
the $|[n,0]|$ iterations, we increment the variable $y$ by $1000$.  We
then non-deterministically execute an inner loop that decrements $y$
by $100$ until $y<100$.  The analysis can derive that only the first
execution of the inner loop depends on the initial value of $y$.  We
again derive a tight bound.

As mentioned, the analysis naturally handles advanced control flow
like \code{break} and \code{return} statements, and mutual recursion.
\pref{fig:ex_rec} contains two mutually-recursive functions with their
automatically derived bounds.  The function \code{count\_down}
decrements its first argument $x$ until it reaches the second argument
$y$. It then recursively calls the function \code{count\_up}.  The
function \code{count\_up} is dual to \code{count\_down}.  Here, we
count up $y$ by $2$ and recursively call \code{count\_down}.  Our
analysis system can compute a tight bound.  This is comparable to the
determination of the point at which to trains meet if they approach
each other with different speeds.

Please not that we can only show a very small selections of the
programs that we can handle automatically here.  In \pref{app:cat} you
find a list of more than $30$ classes of challenging programs that we
can automatically analyze.  We hope that this gives you an impression
of the capabilities of the analysis but we also point out that this
list is far from being exhaustive.

@@

We implemented the new automatic amortized analysis using CIL [XXX].
CIL is a subset of C that is very similar to CompCert Clight but
provides tool support that facilitated the implementation.  The
implementation is publically available at XXX.  More details can be
found in section XXX.

@@

For the sake of simplicity, our quantitative program logic
applies to the Clight language.  This language is the most
abstract intermediate language used by CompCert.  It is very
similar to vanilla C but the expressions are free of
side-effects and there is only one looping construct.
%
\begin{align*}
S &:= \code{assert}~E
\mid \code{skip}
\mid \code{break}
\mid \code{return}~x
\mid x \gets E
\mid x \gets f(x^*)
\\
& \mid \code{loop}~S
\mid \code{if}(E)~S~\code{else}~S
\mid S;S
\end{align*}
%
We added the builtin primitive $\code{assert}$ that checks
if a condition is true before continuing the program
execution.  This will be used to explicit invariants to the
automatic analysis later.

@@

The operational semantics uses a mutable state $\state$
composed of two maps from variable names to integers. The
first map $\theta$ represents the local variable of a
function and the second map $\gamma$ gives values to global
variables of the program.

For simplicity, we assume that local variables and global
variables are always disjoint sets, this simplifies both
the semantics and the description of our analysis.  The
mutable state $\state$ can be accessed and modified, the
access is written functionally and is defined by
$$
(\theta, \gamma)(x) =
\left\{
\begin{array}{lr}
\theta(x) & \mbox{ if } x \in \dom \theta \\
\gamma(x) & \mbox{ if } x \in \dom \gamma
\end{array}
\right..
$$
We note $\state[x \mapsto v]$ the state that results from
updating $\state$ to make $x$ map to $v$.  Similarly to
the state access, the formal definition of this operation
is a simple case disjunction.  Thanks to the assumption we
made above we can be sure that the different cases never
overlap.

To formally describe the cost of operation of a given
program, our semantics tracks resource consumption while
executing.  The semantics is oblivious to the resource of
interest for the user of our system.  We achieve this
independence by parameterizing by a resource
\emph{metric}, it is a tuple of nine rational numbers and
one map from expressions to rational numbers
$$
(
M_a,
M_b,
M_r,
M_u,
M_f,
M_l,
M_c^1,
M_c^2,
M_q,
M_e
).
$$
Each of these rational numbers indicates the amount of resource
consumed by one operation.  If one of these numbers is negative,
it means that resources are released by the program.

For technical reasons in our soundness proof and to be in line
with CompCert's Clight, our language's semantics is defined
in a small-step fashion.  The \pref{fig:opsem} contains the full
list of reduction rules.  These rules provide a rewrite
system for program configurations of the form
$(\state, S, \cont, \cost)$, where $\state$ is the program state,
$S$ is the statement being executed, $\cont$ is a continuation
that describes what remains to be done once the execution of $S$
is over, and $\cost$ is the non-negative quantity of resources
available for further execution.  All rules that can decrease
$\cost$ have the implicit side condition that the resource
quantity available after the step is non-negative.  Using this
resource counter we can define the resource consumption of a
program as the minimal resource counter $\cost_*$ that lets the
program run to termination without errors.

The small-step semantics uses continuations to represent the
context of execution of a given statement.
A continuation can be a sequenced computation, it can also result
from entering a loop or a function, or it can simply be the
empty continuation (used to start a program).
$$
\cont := \Kstop
\mid \Kseq S \cont
\mid \Kloop S \cont
\mid \Kcall r \theta \cont
$$

\begin{figure}[t!]
\begin{mathpar}
%
\Rule{S:Assert}
{ \code{istrue}~\sem{e}_\state }
{ (\state, \code{assert}~e, \cont, \cost)
\smallstep
(\state, \code{skip}, \cont, \cost {-} M_a)
}
%
\and
\Rule{S:BrkSeq}
{}
{ (\state, \code{break}, \Kseq S \cont, \cost)
\smallstep
(\state, \code{break}, \cont, \cost)
}
%
\and
\Rule{S:BrkLoop}
{}
{ (\state, \code{break}, \Kloop \cont, \cost)
\smallstep
(\state, \code{skip}, \cont, \cost {-} M_b)
}
%
\and
\Rule{S:RetSeq}
{}
{ (\state, \code{return}~x, \Kseq S \cont, \cost)
\smallstep
(\state, \code{return}~x, \cont, \cost)
}
\and
%
\and
\Rule{S:RetLoop}
{}
{ (\state, \code{return}~x, \Kloop \cont, \cost)
\smallstep
(\state, \code{return}~x, \cont, \cost)
}
%
\and
\Rule{S:RetCall}
{ \state = (\_, \gamma) \\
  \state' = (\theta, \gamma)[r \mapsto \state(x)]
}
{ (\state, \code{return}~x, \Kcall r \theta \cont, \cost)
\smallstep
(\state', \code{skip}, \cont, \cost {-} M_r)
}
%
\and
\Rule{S:Update}
{ \state' = \state[x \mapsto \sem{e}_\state] }
{ (\state, x \gets e, \cont, \cost)
\smallstep
(\state', \code{skip}, \cont, \cost {-} M_u {-} M_e(e))
}
%
\and
\Rule{S:Call}
{ \delta f = (\vec x, S_f) \\
  \state = (\theta, \gamma) \\
  \state' = (\vec x \mapsto \state(\vec y), \gamma) \\
}
{ (\state, r \gets f(\vec y), \cont, \cost)
\smallstep
(\state', S_f, \Kcall r \theta \cont, \cost {-} M_f)
}
%
\and
\Rule{S:Loop}
{}
{ (\state, \code{loop}~S, \cont, \cost)
\smallstep
(\state, S, \Kloop S \cont, \cost)
}
%
\and
\Rule{S:SkipLoop}
{}
{ (\state, \code{skip}, \Kloop S \cont, \cost)
\smallstep
(\state, \code{loop}~S, \cont, \cost {-} M_l)
}
%
\and
\Rule{S:IfTrue}
{ \code{istrue}~\sem{e}_\sigma }
{ (\state, \code{if}(e)~S_1~\code{else}~S_2, \cont, \cost)
\smallstep
(\state, S_1, \cont, \cost {-} M_c^1)
}
%
\and
\Rule{S:IfFalse}
{ \code{isfalse}~\sem{e}_\sigma }
{ (\state, \code{if}(e)~S_1~\code{else}~S_2, \cont, \cost)
\smallstep
(\state, S_2, \cont, \cost {-} M_c^2)
}
%
\and
\Rule{S:Seq}
{ }
{ (\state, S_1; S_2, \cont, \cost)
\smallstep
(\state, S_1, \Kseq {S_2} \cont, \cost)
}
%
\and
\Rule{S:SkipSeq}
{}
{ (\state, \code{skip}, \Kseq S \cont, \cost)
\smallstep
(\state, S, \cont, \cost {-} M_q)
}
%

@@

@@

\begin{figure*}[t]
\begin{mathpar}
%
\Rule{L:Assert}
{ }
{ \Delta; B; R \vdash
\htriple
  { Q \implies \code{istrue}~\sem{e}_\state }
  { \code{assert}~e }
  { Q }
}
%
\and
\Rule{L:Skip}
{ }
{ \Delta; B; R \vdash
\htriple
  { M_s + Q }{ \code{skip} }{ Q }
}
%
\and
\Rule{L:Break}
{ }
{ \Delta; B; R \vdash
\htriple
  { M_b + B }{ \code{break} }{ Q }
}
%
\and
\Rule{L:Return}
{ }
{ \Delta; B; R \vdash
\htriple
  { M_r + R\,\sem{e}_\state }{ \code{return}~e }{ Q }
}
%
\and
\Rule{L:Update}
{ }
{ \Delta; B; R \vdash
\htriple
  { \lambda \state.\,M_u + Q\,\state[x \mapsto \sem{e}_\state] }{ x \gets e }{ Q }
}
%
\and
\Rule{L:Call}
{
  \Delta f = \forall z\,\vec v.(P_f\,z\,\vec v, Q_f\,z) \\
  P \models P_f \, y \, (\state(\vec x)) \land A \\
  \forall v.\, (Q_f \, y \, v \land A \models \lambda \state.\, Q \, \state[r \mapsto v])
}
{ \Delta; B; R \vdash
\htriple
  { M_f + P }
  { r \gets f(\vec x) }
  { Q }
}
%
\and
\Rule{L:Loop}
{ \Delta; Q; R \vdash
\htriple
  { I }{ S }{ I + M_l }
}
{ \Delta; B; R \vdash
\htriple
  { I }{ \code{loop}~S }{ Q }
}
%
\and
\Rule{L:Seq}
{ \Delta; B; R \vdash
\htriple{ P }{ S_1 }{ Q' + M_s }
\\
  \Delta; B; R \vdash
\htriple{ Q' }{ S_2 }{ Q }
}
{ \Delta; B; R \vdash
\htriple{ P }{ S_1;S_2 }{ Q }
}
%
\and
\Rule{L:If}
{ \Delta; B; R \vdash
\htriple
  { P \land \code{istrue}~\sem{e}_\state - M_c^1}
  { S_1 }{ Q }
\\
  \Delta; B; R \vdash
\htriple
  { P \land \code{isfalse}~\sem{e}_\state - M_c^2 }
  { S_2 }{ Q }
}
{ \Delta; B; R \vdash
\htriple
  { P }{ \code{if}(e)~S_1~\code{else}~S_2 }{ Q }
}
%
\and
\Rule{L:Weaken}
{
  P \models P' \\
  \Delta; B'; R' \vdash
  \htriple{ P' }{ S }{ Q' } \\
  Q' \models Q \\
  B' \models B \\
  \forall v.\, (R'\,v \models R\,v)
}
{ \Delta; B; R \vdash
\htriple{ P }{ S }{ Q }
}
%
\and
\Rule{L:Frame}
{ \Delta; B; R \vdash
\htriple{ P }{ S }{ Q } \\
  x \in \mathbb Q_0^+
}
{ \Delta; B + x; R + x \vdash
\htriple{ P + x }{ S }{ Q + x }
}
%
\and
\Rule{L:Extend}
{ \forall f\,P_f\,Q_f.\,
  \Delta'f = \forall z\,\vec v.(P_f\,z\,\vec v, Q_f\,z)
  \rightarrow
  \forall y\,\vec v.\,
  (\Delta \cup \Delta'; \bot; Q_f\,y \vdash
    \htriple{ P_f\,y\,\vec v }{ S_f }{ \bot })
\\
  \Delta \cup \Delta'; B; R \vdash
  \htriple{ P }{ S }{ Q }
}
{ \Delta; B; R \vdash
\htriple{ P }{ S }{ Q }
}
%
\end{mathpar}
